# Visualization

In this section we'll explore visualization methods in R. Visualization has been a key element of R since its inception, since visualization is central to the exploratory philosophy of the language. The base *plot* system generally does a good job in coming up with the most likely graphical output based on the data you provide. 

```{r}
plot(iris$Petal.Length, iris$Sepal.Length)
plot(iris$Species, iris$Petal.Length)
```

## Introducing ggplot2

We'll mostly focus however on gpplot2, based on the *Grammar of Graphics* because it provides considerable control over your graphics while remaining fairly easily readable, as long as you buy into its grammar.

ggplot2 looks at three aspects of a graph:

- data : where are the data coming from?
- geometry : what type of graph are we creating?
- aesthetics : what choices can we make about symbology and how do we connect symbology to data?

See https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

The ggplot2 system provides plots of single and multiple variables, using various coordinate systems (including geographic).

```{r include=FALSE}
library(tidyverse)
XSptsNoTrees <- read_csv("data/XSptsNDVI.csv") %>%
  filter(vegetation != "pine")
XSptsPheno <- XSptsNoTrees %>%
  pivot_longer(cols = starts_with("NDVI"), names_to = "phenology", values_to = "NDVI") %>%
  mutate(phenology = str_sub(phenology, 5, str_length(phenology)))
write_csv(XSptsPheno, "data/XSptsPheno.csv")

```

## Plotting one variable

- continuous
    - histograms
    - density plots
    - dot plots
- discrete
    - bar
    
```{r}
ggplot(XSptsPheno, aes(vegetation)) + 
   geom_bar()

```

### Histogram

```{r warning=FALSE}
XSptsPheno <- read_csv("data/XSptsPheno.csv")
XSptsPheno %>%
  ggplot(aes(NDVI)) + 
  geom_histogram(binwidth=0.05)
```

**Normal histogram**: easier to visualize the distribution, see modes
```{r warning=FALSE}
sierraData <- read_csv("data/sierraData.csv")
sierraData %>%
    ggplot(aes(TEMPERATURE)) +
  geom_histogram(fill="dark green")
```

**Cumulative histogram with proportions**: easier to see percentiles, median
```{r warning=FALSE}
n <- length(sierraData$TEMPERATURE)
sierraData %>%
  ggplot(aes(TEMPERATURE)) +
  geom_histogram(aes(y=cumsum(..count..)/n), fill="dark goldenrod")

```

### Density Plot

Density represents how much out of the total. The total area (sum of widths of bins times densities of that bin) adds up to 1.

```{r}
XSptsPheno %>% 
  ggplot(aes(NDVI)) + 
  geom_density()

```

Note that NDVI values are <1 so bins are very small numbers, so in this case densities can be >1.

**Using alpha and mapping phenology as fill color**.  This illustrates two useful ggplot methods:

- "mapping" a variable (phenology) to an aesthetic property (fill color of the density polygon)
- setting a a property (alpha = 0.2) to all polygons of the density plot.  The alpha channel of colors defines its opacity, from invisible (0) to opaque (1) so is commonly used to set as its reverse, transparency.

```{r}
XSptsPheno %>%
  ggplot(aes(NDVI, fill=phenology)) +
  geom_density(alpha=0.2)
```

```{r include=FALSE}
library(tidyverse)
runoffPivot <- read_csv("data/eucoakrainfallrunoffTDR.csv") %>%
  pivot_longer(cols = starts_with("runoffL_"), 
               names_to = "tree", values_to = "runoff_L") %>%
  mutate(
    tree = str_sub(tree, str_length(tree)-2, str_length(tree)),
    Date = as.Date(date, "%m/%d/%Y"))  # variable date is just a string, needs converting to date format

euc <- runoffPivot %>%
  filter(tree == "euc") %>%
  mutate(
    rain_subcanopy = rain_euc,
    slope = slope_euc,
    aspect = aspect_euc,
    surface_tension = surface_tension_euc,
    runoff_rainfall_ratio = runoff_rainfall_ratio_euc) %>%
  select(site, `site #`, tree, Date, month, rain_mm, rain_subcanopy, slope, aspect, runoff_L, surface_tension, runoff_rainfall_ratio)

oak <- runoffPivot %>%
  filter(tree == "oak") %>%
  mutate(
    rain_subcanopy = rain_oak,
    slope = slope_oak,
    aspect = aspect_oak,
    surface_tension = surface_tension_oak,
    runoff_rainfall_ratio = runoff_rainfall_ratio_oak) %>%
  select(site, `site #`, tree, Date, month, rain_mm, rain_subcanopy, slope, aspect, runoff_L, surface_tension, runoff_rainfall_ratio)

eucoak <- rbind(euc, oak)
write_csv(eucoak, "data/tidy_eucoak.csv")

```

```{r warning=FALSE}
runoffEucOak <- read_csv("data/tidy_eucoak.csv")
runoffEucOak %>%
  ggplot(aes(log(runoff_L),fill=tree)) +
  geom_density(alpha=0.2)
```


### boxplot

```{r warning=FALSE}
ggplot(data = runoffEucOak) +
  geom_boxplot(aes(x = site, y = runoff_L))

```
**Get color from tree within `aes()`**
```{r warning=FALSE}
ggplot(data = runoffEucOak) +
  geom_boxplot(aes(x=site, y=runoff_L, color=tree))

```

**Visualizing soil CO_2_ data with a Tukey box plot**

```{r warning=FALSE}
co2 <- read_csv("data/soilCO2_97.csv")
co2$SITE <- factor(co2$SITE)  # in order to make the numeric field a factor
ggplot(data = co2, mapping = aes(x = SITE, y = `CO2%`)) +
  geom_boxplot()

```

## Plotting two variables
### Two continuous variables

We've looked at this before -- the scatterplot

```{r warning=FALSE}
sierra <- read_csv("data/sierraFeb.csv")
ggplot(data=sierra) + 
  geom_point(mapping = aes(TEMPERATURE, ELEVATION))

```

- The aes ("aesthetics") function specifies the variables to use as x and y coordinates 
- geom_point creates a scatter plot of those coordinate points

**Set color for all (*not* in aes())**

```{r warning=FALSE}
ggplot(data=sierra) + 
  geom_point(aes(TEMPERATURE, ELEVATION), color="blue")
```

- color is defined outside of aes, so is applies to all points.
- mapping is first argument of  geom_point, so `mapping = ` is not needed.

### Two variables, one discrete

```{r warning=FALSE}
ggplot(runoffEucOak) +
  geom_bar(aes(site, runoff_L), stat="identity")

```

## Color systems

You can find a lot about color systems. See these sources:

<a href="http://sape.inf.usi.ch/quick-reference/ggplot2/colour">http://sape.inf.usi.ch/quick-reference/ggplot2/colour</a>
<a href="http://applied-r.com/rcolorbrewer-palettes/">http://applied-r.com/rcolorbrewer-palettes/</a>

### Color from variable, in aesthetics

In this graph, color is defined inside aes, so is based on COUNTY

```{r warning=FALSE}
sierra <- read_csv("data/sierraFeb.csv")
ggplot(data=sierra) + 
  geom_point(aes(TEMPERATURE, ELEVATION, color=COUNTY))

```

**Plotting lines using the same x,y in aesthetics**
```{r warning=FALSE}
sierra %>%
  ggplot(aes(TEMPERATURE,ELEVATION)) +
  geom_point(color="blue") +
  geom_line(color="red")
```

Note the use of pipe to start with the data then apply ggplot.

**River map & profile**

```{r}
x <- c(1000, 1100, 1300, 1500, 1600, 1800, 1900)
y <- c(500, 700, 800, 1000, 1200, 1300, 1500)
z <- c(0, 1, 2, 5, 25, 75, 150)
d <- rep(NA, length(x))
longd <- rep(NA, length(x))
s <- rep(NA, length(x))
for(i in 1:length(x)){
  if(i==1){longd[i] <- 0; d[i] <-0}
  else{
    d[i] <- sqrt((x[i]-x[i-1])^2 + (y[i]-y[i-1])^2)
    longd[i] <- longd[i-1] + d[i]
    s[i-1] <- (z[i]-z[i-1])/d[i]}}
longprofile <- bind_cols(x=x,y=y,z=z,d=d,longd=longd,s=s)
ggplot(longprofile, aes(x,y)) +
  geom_line(mapping=aes(col=s), size=1.2) + 
  geom_point(mapping=aes(col=s, size=z)) +
  coord_fixed(ratio=1) + scale_color_gradient(low="green", high="red") +
  ggtitle("Simulated river path, elevations and slopes")
ggplot(longprofile, aes(longd,z)) + geom_line(aes(col=s), size=1.5) + geom_point()  +
  scale_color_gradient(low="green", high="red") +
  ggtitle("Elevation over longitudinal distance upstream")
ggplot(longprofile, aes(longd,s)) + geom_point(aes(col=s), size=3) +
  scale_color_gradient(low="green", high="red") +
  ggtitle("Slope over longitudinal distance upstream")
#summary(lm(s~longd, data=longprofile))

```

### Trend line

```{r}
sierra %>%
  ggplot(aes(TEMPERATURE,ELEVATION)) +
  geom_point(color="blue") +
  geom_smooth(color="red", method="lm")

```

### General symbology
A useful vignette lets you see aesthetic specifications for symbols, including:

- Color & fill
- Lines
   - line type, size, ends
- Polygon
   - border color, linetype, size
   - fill
- Points
   - shape
   - size
   - color & fill
   - stroke
- Text
   - font face & size
   - justification

```{r}
vignette("ggplot2-specs")
```

#### Categorical symbology


EPA Toxic Release Inventory.  A big data set needing symbology to try to make sense of it.

```{r message=FALSE}
TRI <- read_csv("data/TRI_2017_CA.csv") %>%
  filter(`5.1_FUGITIVE_AIR` > 100 & `5.2_STACK_AIR` > 100)
ggplot(data = TRI, aes(log(`5.2_STACK_AIR`), log(`5.1_FUGITIVE_AIR`), 
                       color = INDUSTRY_SECTOR)) +
       geom_point()
```

#### Graphs from grouped data

```{r warning=FALSE}
XSptsPheno %>%
  ggplot() +
  geom_point(aes(elevation, NDVI, shape=vegetation, 
                 color = phenology), size = 3) +
  geom_smooth(aes(elevation, NDVI, 
                 color = phenology), method="lm") 
```

```{r warning=FALSE}
ggplot(data = eucoak) +
  geom_point(mapping = aes(x = rain_mm, y = runoff_L, color = tree)) +
  geom_smooth(mapping = aes(x = rain_mm, y= runoff_L, color = tree), 
              method = "lm") +
  scale_color_manual(values = c("seagreen4", "orange3"))

```

#### Faceted graphs

This is another option to displaying groups of data, with parallel graphs

```{r warning=FALSE}
ggplot(data = runoffEucOak) +
  geom_point(aes(x=rain_mm,y=runoff_L)) +
  geom_smooth(aes(x=rain_mm,y=runoff_L), method="lm") +
  facet_grid(tree ~ .)

```

## Titles and subtitles

```{r warning=FALSE}
ggplot(data = runoffEucOak) +
  geom_point(aes(x=rain_mm,y=runoff_L, color=tree)) +
  geom_smooth(aes(x=rain_mm,y=runoff_L, color=tree), method="lm") +
  scale_color_manual(values=c("seagreen4","orange3")) +
  labs(title="rainfall ~ runoff", 
       subtitle="eucalyptus & oak sites, 2016")

```

## Pairs Plot

```{r warning=FALSE}
sierra %>%
  select(LATITUDE, ELEVATION, TEMPERATURE, PRECIPITATION) %>%
  pairs()
```
There are many versions of pairs plots. Here's one from GGally (need to `install.packages("GGally")` first):

```{r warning=FALSE}
library(GGally)
sierra %>%
   select(LATITUDE, ELEVATION, TEMPERATURE, PRECIPITATION) %>%
   ggpairs()

```


