# Data Transformation

The goal of this section is to continue where we started in the earlier chapter on data abstraction with **`dplyr`** to look at more transformational functions, and **`tidyr`** adds other tools like pivot tables.

- **`dplyr`** tools: 
   - joins: `left_join`, `right_join`, `inner_join`, `full_join`, `semi_join`, `anti_join`
   - set operations: `intersect`, `union`, `setdiff`
   - binding rows and columns: `bind_cols`, `bind_rows`
- **`tidyr`** tools:
   - pivot tables: `pivot_longer`, `pivot_wider`
   

The term "data wrangling" has been used for what we're doing with these tools, and the
relevant cheat sheet is actually called "Data Wrangling"
<a href="https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf">https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf</a>

## Data joins

To bring in variables from another data frame based on a common join field.
There are multiple types of joins. Probably the most common is **`left_join`** since it starts
from the data frame (or sf) you want to continue working with and bring in data
from an additional source. You'll retain all records of the first data set. 
For any non-matches, NA is assigned.

```{r message=F, warning=F}
library(tidyverse)
library(iGIScData)
library(sf)
csvPath <- system.file("extdata", "CA_MdInc.csv", package = "iGIScData")
income <- read_csv(csvPath) %>%
   select(trID, HHinc2016) %>%
   mutate(HHinc2016 = as.numeric(HHinc2016),
          joinid = str_c("0", trID)) %>%
   select(joinid, HHinc2016)
census <- BayAreaTracts %>%
   left_join(income, by = c("FIPS" = "joinid")) %>%
   select(FIPS, POP12_SQMI, POP2012, HHinc2016)
head(census %>% st_set_geometry(NULL))
```

Other joins are:

- **`right_join`** where you end up retaining all the rows of the second data set
and NA is assigned to non-matches
- **`inner_join`** where you only retain records for matches
- **`full_join`** where records are retained for both sides, and NAs assigned to non-matches

**Right join example**
We need to join NCDC monthly climate data for all California weather stations to a selection of 82 stations that are in the Sierra. 

- The monthly data has 12 rows (1/month) for each station
- The right_join gets all months for all stations, so we weed out the non-Sierra stations by removing NAs from a field only with Sierra station data

```{r message=F}
sierra <- right_join(sierraStations, CA_ClimateNormals, by="STATION") %>%
   filter(!is.na(STATION_NA)) %>% select(-STATION_NA)
head(sierra %>% filter(DATE == "01") %>% select(NAME, ELEVATION, `MLY-TAVG-NORMAL`), n=10)
```

The exact same thing however could be accomplished with an inner_join and
it doesn't required removing the NAs:

```{r message=F}
sierraAlso <- inner_join(sierraStations, CA_ClimateNormals, by="STATION") %>%
   select(-STATION_NA)
```

## Set Operations

Set operations compare two data frames (or vectors) to handle observations or rows that 
are the same for each, or not the same. The three set methods are:

- `dplyr::`**`intersect(x,y)`** retains rows that appear in *both* x and y
- `dplyr::`**`union(x,y)`** retains rows that appear in either or both of x and y
- `dplyr::`**`setdiff(x,y)`** retains rows that appear in x but not in y

```{r message=F}
squares <- (1:10)^2
evens <- seq(0,100,2)
squares
evens
intersect(squares,evens)
sort(union(squares,evens))
sort(setdiff(squares,evens))
```

## Binding Rows and Columns

These `dplyr` functions are similar to `cbind` and `rbind` in base R, but always 
creates data frames. For instance, `cbind` usually creates matrices, and make all vectors
the same class. Note that in `bind_cols`, the order of data in rows must be the same.

```{r message=F}
states <- bind_cols(abb=state.abb,
                    name=state.name,
                    region=state.region,
                    state.x77)
head(states)
```

To compare, note that `cbind` converts numeric fields to character when any other 
field is character, and character fields are converted to character integers where there are any repeats, 
which would require manipulating them into factors:

```{r message=F}
states <- as_tibble(cbind(abb=state.abb, 
                          name=state.name, 
                          region=state.region,
                          division=state.division,
                          state.x77))
head(states)
```

## Pivotting data frames

Pivot tables are a popular tool in Excel, allowing you to transform your data to be more
useful in a particular analysis. A common need to pivot is 2+ variables with the same data where the variable name should be a factor. `Tidyr` has **`pivot_wider`** and **`pivot_longer`**.  

- **`pivot_wider`** pivots rows into variables.
- **`pivot_longer`** pivots variables into rows, creating factors.

In our meadows study cross-section created by intersecting normalized difference vegetation
index (NDVI) values from multispectral drone imagery with surveyed elevation and vegetation
types (xeric, mesic, and hydric), we have fields `NDVIgrowing` from a July 2019 growing season and `NDVIsenescent` from a September 2020 dry season, but would like 'growing' and 'senescent' to be factors with a single `NDVI` variable. This is how we used `pivot_longer` to accomplish this, using data
from the `iGIScData` data package:

```{r message=F}
XSptsPheno <- XSptsNDVI %>%
      pivot_longer(cols = starts_with("NDVI"), 
                   names_to = "phenology", values_to = "NDVI") %>%
      mutate(phenology = str_sub(phenology, 5, str_length(phenology)))
```

Then to do the opposite use `pivot_wider`:

```{r message=F}
XSptsPheno %>%
  pivot_wider(names_from = phenology, names_prefix = "NDVI", 
              values_from = NDVI)
XSptsPheno
XSptsPheno %>%
  ggplot() +
  geom_point(aes(elevation, NDVI, shape=vegetation, 
                 color = phenology), size = 5) +
  geom_smooth(aes(elevation, NDVI, 
                 color = phenology), method="lm")
```

Pivots turn out to be commonly useful. Runoff graphing from the Eucalyptus/Oak study also benefitted
from a pivot_longer:

```{r message=F, warning=F}
eucoakrainfallrunoffTDR %>%
  pivot_longer(cols = starts_with("runoffL"),
               names_to = "tree", values_to = "runoffL") %>%
  mutate(tree = str_sub(tree, str_length(tree)-2, str_length(tree))) %>%
  ggplot() + geom_boxplot(aes(site, runoffL)) +
    facet_grid(tree ~ .)
```



