---
title: "Intro to Shiny: Input/Output"
author: "Jerry Davis"
date: "4/17/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Shiny lets you include interactive controls letting the user manipulating your R analysis.

Since Shiny requires a host to run the R code on, we can't create it as part of 
a chapter in the book. But we can look at it in this RMarkdown document, which can run
RMarkdown using the 'shiny' runtime method in RStudio.  We'll explore Shiny interactive input widgets and
output functions, then look at using these to build a Shiny app which can be 
run on your computer or hosted on the web.

To learn more about Shiny apps, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

## Inputs and Outputs

We'll start by looking at Shiny inputs and outputs:  

- Inputs are the interactive controls (widgets) that let the user change the resulting output.
- Outputs are the graphs, maps or tables, and are automatically updated whenever inputs change. 

The following includes a couple of widget-controlled input settings to be used in a plot produced by `renderPlot`

- a `selectInput` that is used to set the `breaks` parameter for the `hist` function as a number of bins
- a `sliderInput` to set the `adjust` (bandwidth adjustment) parameter for the density plot

Note how the input variables are accessed by the output function as `input$n_breaks` and `input$bw_adjust`:

```{r eruptions}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

## Input widgets

You can explore the various input and output methods in the cheat sheet...

<a href="https://shiny.rstudio.com/images/shiny-cheatsheet.pdf">https://shiny.rstudio.com/images/shiny-cheatsheet.pdf</a>

... but we'll start by looking at a selection on input widgets and see what they produce.
Here's a couple of simple ones, with a simple print output:

```{r}
numericInput(inputId="n", "sample size", value = 25)
renderPrint(print(input$n))
sliderInput(inputId="i", "input", min=1, max=30, value = 15)
renderPrint(print(input$i))
```

### Creating a plot

The above widgets went to a rendered print output, but the same simple inputs can of course be used to create a plot:

```{r}
sliderInput(inputId="bins", "number of bins", min=1, max=30, value = 15)
renderPlot(hist(rnorm(100), breaks=input$bins))
```

### Other input widgets

There are lots of other input widgets that are pretty easy to see how they apply based on the type of control we need to set:

- **`radioButtons()`** for choosing just one, and by default the first is chosen
```{r}
radioButtons(inputId="which_one", label="Select:", choices = c("Choice 1"="Choice1","Choice 2"="Choice2","Choice 3"="Choice3"))
renderPrint(print(input$which_one))
```

- **`checkboxGroupInput()`** for choosing multiples
```{r}
checkboxGroupInput(inputId="which", label="Select:", choices = c("Choice 1"="Choice1","Choice 2"="Choice2","Choice 3"="Choice3"))
renderPrint(print(input$which))
```

- **`dateInput()`**
```{r}
dateInput(inputId="date", label="Select date:")
renderPrint(print(input$date))
```

- **`fileInput()`**
```{r}
fileInput(inputId="file", label="Select file:")
renderPrint(print(input$file))
```

- **`textInput()`**
```{r}
textInput(inputId="text", label="Enter text:")
renderPrint(print(input$text))
```

- **`actionLink()`**
```{r}
actionLink(inputId="action", label="Action:")
renderPrint(print(input$action))
```

