---
title: "Shiny Apps"
author: "Jerry Davis"
date: "4/17/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this document, we'll explore the very basics of Shiny Web Apps, with complete code that
can be used to create the app.R script from which it's run. Shiny apps can be created in RStudio as a *Shiny Web Application* project type. 

## Shiny Apps (created inline for viewing here)

This R Markdown document is made interactive using Shiny, and employs a inline applications
with the complete Shiny app code included, so we can explore various input and output methods. All code shown can also be used to create a Shiny web app by saving it as `app.R` in an RStudio Shiny app project.

### A simple Shiny app

The following simple Shiny app illustrates the basic elements:

- Optionally some code at the top to set things up, such as loading packages and setting up data.
- The **ui** section, which determines what the user will see, such as input widgets (like `numericInput`) and an area to plot things in, in this case `plotOutput`. In order to fit well on the page, the `fluidPage` function is used to hold it in; we could have also used `inputPanel` as we did earlier, but the result is not as well laid out.
- The **server** section, which uses two parameters (input, output) created in the user interface to:
   - provide data from the input widget (in this case `numericInput` which creates `input$n`) such as parameter settings to functions (e.g. `rnorm()` here which needs a sample size provided by `input$n`) in the rendered output (of in this case `hist`) 
   - to render an output using a render___ method such as `renderPlot` which will run everything inside it (in this case a histogram)
- Providing that `ui` and `server` to the `shinyApp` function.

```{r simpleInline, warning=F, message=F}
library(shiny)
ui <- fluidPage(
    numericInput(input="n",
                 "Sample size", value=25),
    plotOutput(outputId = "hist"))
server <- function(input, output) {
    output$hist <- renderPlot({
        hist(rnorm(input$n))
    })
}   
shinyApp(ui=ui,server=server)
```

This Shiny app is very short, which is useful in wrapping your head around these
basic elements, which is important to understand. You can use this exact code to 
create `app.R` in a Shiny web application project to see what you get when you click
the Run App button. 

- Create a new RStudio instance, and create a new Shiny Web Application. It will create the Old Faithful
app.  Have a look at the code and use Run App to see what it does.
- Then replace the app.R code with the above even simpler code, and check it out.

## Shiny app I/O methods

One thing to note about how the Shiny web app (or inline or embedded app) works as opposed
to just using the Shiny I/O methods we looked at earlier is the way the ui and server communicate
about the plot creation: We didn't use `plotOutput` in the I/O methods; we just used 
the input widgets to set parameters in the `renderPlot`.

Instead, this Shiny web app has the `renderPlot` in the **server** section, and a 
new function `plotOutput` is used to both specify the location of the plot and set
`outputId` to be used as a variable to append to `output$` in the **server** section.
Referring to the **Outputs** section of the cheat sheet, we can see that various render\*() 
functions work with corresponding \*Output() functions:

- `renderDataTable()` works with `dataTableOutput()`
- `renderImage(    )` works with `imageOutput()`
- `renderPlot(     )` works with `plotOutput()`
- `renderPrint(    )` works with `verbatimTextOutput()`
- `renderTable(    )` works with `tableOutput()`
- `renderText(     )` works with `textOutput()`
- `renderUI(       )` works with `uiOutput()` and `htmlOutput()`

And there are others.  For instance, as we'll see in the next document:

- `renderLeaflet(  )` works with `leafletOutput()`


### Data Tables:  `renderDataTable()` and `dataTableOutput()`

```{r DataTable, warning=F, message=F}
library(shiny); library(tidyverse)
ui <- fluidPage(
    numericInput(input="n",
                 "Sample size", value=25),
    dataTableOutput(outputId = "varTable"))
server <- function(input, output) {
    output$varTable <- renderDataTable(tibble(a=rnorm(input$n), b=rnorm(input$n)))}
shinyApp(ui=ui,server=server)
```


### Text as Character:  `renderPrint()` and `verbatimTextOutput()`

These functions produce a mono-spaced-font output like a character variable.

```{r PrintOut, warning=F, message=F}
library(shiny); library(tidyverse)
ui <- fluidPage(
    textInput(inputId="monotxt", label="Enter text:", value="Some text"),
    verbatimTextOutput(outputId = "txt"))
server <- function(input, output) {
    output$txt <- renderPrint(print(input$monotxt))}
shinyApp(ui=ui,server=server)
```


### Formatted text: `renderText()` and `textOutput()`

```{r TextOut, warning=F, message=F}
library(shiny); library(tidyverse)
ui <- fluidPage(
    textInput(inputId="text", label="Enter text:", value="Some text"),
    textOutput(outputId = "prn"))
server <- function(input, output) {
    output$prn <- renderText(print(input$text))}
shinyApp(ui=ui,server=server)
```

### Plots: `renderPlot()` and `plotOutput()`

```{r PlotOut, warning=F, message=F}
library(shiny); library(tidyverse)
ui <- fluidPage(
    sliderInput(inputId="i", "input", min=1, max=30, value = 15),
    textInput(inputId="title", label="Enter title:", value="Tukey boxplot"),
    plotOutput(outputId = "Tukey"))
server <- function(input, output) {
    output$Tukey <- renderPlot(boxplot(rnorm(input$i),main=input$title))}
shinyApp(ui=ui,server=server)
```


## Learn more:

You should explore these resources to learn more about creating useful Shiny web apps:

- <a href="https://shiny.rstudio.com/images/shiny-cheatsheet.pdf">https://shiny.rstudio.com/images/shiny-cheatsheet.pdf</a>
- The Shiny tutorial at <a href="https://shiny.rstudio.com/tutorial/">https://shiny.rstudio.com/tutorial/</a>, worth spending some time on to learn more about getting Shiny apps working. There's a lot to understand about reactivity for instance. 


To learn more about inline apps in R Markdown, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).




