---
title: "Shiny App Example: MODIS fires"
author: "Jerry Davis"
date: "4/17/2021"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Shiny app as Inline Application

This R Markdown document is made interactive using Shiny, and employs an Inline Application with the complete Shiny app code included. 

The code shown can also be used to create a Shiny web app by saving it as `app.R` in an RStudio Shiny app project, though you will need the `"Shiny/data"` folders set up to store data; these are used to keep the location with Shiny Rmd files uncluttered with the downloaded shapefile data. In terms of complexity of inputs, this app is pretty straightforward -- the fluidPage is pretty standard for the ui section.  The leaflet map is pretty straightforward (though getting the point symbology right took some experimentation.  There are also some other tricks that make it a bit more complicated, such as web scraping to download new data, and using `observe` as a way to observe some parameters to allow the current zoom level to stick.

*Note that it takes a while for the map to appear because the data is being downloaded and processed.*
*Every time you change to a new year, there is also a delay as new data is downloaded and processed.*

```{r MODISfires, warning=F, message=F, collapse=T}
library(sf); library(leaflet); library(tidyverse); library(lubridate)
library(shiny); library(here)
setwd(paste(here::here(),"/Shiny/data",sep=""))
dataPath = getwd()
pal <- colorNumeric(c("orange", "firebrick4"),domain=300:550)
ui <- fluidPage(
    titlePanel("MODIS fire detections from temperature anomalies"), 
    leafletOutput("view"),
    sliderInput(inputId = "end_jday",
                label = "ending Julian day",
                value = yday(now()), min = 1, max = 365, step = 1),
    sliderInput(inputId = "numdays",
                label = "number of days",
                value = 3, min = 1, max = 14, step = 1),
    sliderInput(inputId = "year",
                label = "year",
                value = year(now()), min = 2009, max = year(now()), step = 1, sep = ""),
    helpText("Jerry Davis, SFSU Institute for Geographic Information Science. Data source: USDA Forest Service https://fsapps.nwcg.gov/afm/gisdata.php")
)

server <- function(input, output, session) {
    output$view <- renderLeaflet({   # Only year is reactive, so only runs initially or with year change
        yrst <- as.character(input$year)
        txt <- read_file(str_c("https://fsapps.nwcg.gov/afm/data/fireptdata/modisfire_",yrst,"_conus.htm"))
        shpPath <- str_extract(txt, "https://fsapps.nwcg.gov/afm/data/fireptdata/modis_fire_.........conus_shapefile.zip")
        shpZip <- str_extract(shpPath, "modis_fire_.........conus_shapefile.zip")
        MODISfile <- str_c(dataPath,"/",str_extract(shpZip, "modis_fire_.........conus"),".shp")
        if(yrst == as.character(year(now())) | !file.exists(MODISfile)) {
            shpZipPath <- str_c(dataPath, "/",shpZip)
            download.file(shpPath, shpZipPath)
            unzip(shpZipPath, exdir=dataPath) }
        fires <<- st_read(MODISfile)       
        leaflet() %>%
            addProviderTiles(providers$Esri.WorldTopoMap) %>%
            fitBounds(-123,37,-120,39)
    })
    observe({                   # Allows the map to retain its location and zoom
        numdays <- input$numdays; end_jday <- input$end_jday
        fireFilt <- filter(fires,between(JULIAN, end_jday - numdays, end_jday))
        yrst <- as.character(input$year)
        dat <- as.Date(end_jday-1, origin=str_c(yrst,"-01-01"))  # Have to set Julian day one less to relate to Jan 1
        leafletProxy("view", data = fireFilt) %>%
            clearMarkers() %>%
            addCircleMarkers(
                radius = ~(TEMP-250)/50,  # scales 300-500 from 1:5 
                color = ~pal(TEMP),
                stroke = FALSE, fillOpacity = 0.8) %>%
            clearControls() %>%   # clears the legend
            addLegend("topright", pal=pal, values=~TEMP, opacity=0.6,
                title=str_c("MODIS °K</br>",numdays, " days: ", dat,"</br>maxJulian: ",as.character(max(fires$JULIAN))))
    })
}
shinyApp(ui = ui, server = server, options = list(width = "100%", height = 800))
```

To learn more about interactive documents in R Markdown, see [Interactive Documents](http://rmarkdown.rstudio.com/authoring_shiny.html).

Some good resources on Shiny apps:

- <a href="https://shiny.rstudio.com/images/shiny-cheatsheet.pdf">https://shiny.rstudio.com/images/shiny-cheatsheet.pdf</a>
- The Shiny tutorial at <a href="https://shiny.rstudio.com/tutorial/">https://shiny.rstudio.com/tutorial/</a>, worth spending some time on to learn more about getting Shiny apps working. There's a lot to understand about reactivity for instance. 



